From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions

### 2008-09-01 thesis/paper.tex
Preliminary final version of theory

\documentclass[a4paper]{article}

%\usepackage[english]{babel}

@@

\usepackage{fontspec}
\usepackage[xetex,colorlinks]{hyperref}
\begin{document}
\lstset{language=Prolog,numbers=left,numberstyle=\tiny,numbersep=5pt,stepnumber=2,basicstyle=\ttfamily,identifierstyle=\textbf,keywordstyle=\underbar}
\renewcommand{\thepage}{\roman{page}}
\maketitle
\thispagestyle{empty}
\begin{center}
  \vskip 4em
  Thesis submitted for the Degree of Bachelor of Arts\\

  \paragraph{Supervisor:} Dr. Frank Richter
  \vskip 6em
  Seminar für Sprachwissenschaft\\
  Eberhard-Karls-Universität\\
  Tübingen
\end{center}
\newpage
\tableofcontents
\newpage
\SpruechleAufsagen{Aleksandar L. Dimitrov}
\newpage
\renewcommand{\thepage}{\arabic{page}}
\setcounter{page}{1}
\section{Introduction}

@@

\section{Theory}\label{sec:theory}

@@

\subsection{Formalism: Models}\label{sec:formal}

@@

As was already mentioned in \ref{sec:curt}, \curt{} is based on 
the use of external tools. These tools are only capable of treating first order logic – an
extensional framework. \prol{} on the other hand, is a language that can very well
make use of higher order formalisms for calculation\footnote{The two most
popular predicates in \prol{} used for implementing higher-order logic are
\code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog} which
provides an implementation of the former, but not of the latter. In addition
\pn{SWI-Prolog} also defines several higher order aggregate functions, including
\code{maplist/4}, which will be used later on. See \cite{naish:prolhio} for a
discussion of higher order functional programming in \prol.}.  While there exist
inferencing tools for modal logic\footnote{ The interested reader is referred to
\pn{Molle} (\url{http://molle.sf.net}), a modal theorem prover that comes with a
graphical user interface which allows interactive investigation of the
constructed indices.}, a restructuring of the \curt{} system to take advantage
of them would be beyond the scope of this thesis.  It was therefore decided to
avoid modal logic and implement a pragmatic approach to representing indices in
order to increase the expressive power of the system.

Some research on the logic of questions (sometimes also referred to as eroteric
logic) suggests that the treatment of interrogatives demands its own formalism.
\acurt{} employs a formalism that departs from both \abbr{IL} and \abbr{Ty2} in
some ways. Most notably, the notion of a \term{model} is changed as follows.

@@

One of the more important differences to the `traditional', \abbr{FOL}, notion of a model, is
that it was decided to assign every world a unique domain of discourse referents
or \term{individuals}. Note that there is no requirement $\pair{i,\pair{D,F}},
\pair{j,\pair{D',F'}} \in M \wedge D\neq D'$, and most of the time, a model's
possible worlds will indeed share many of their individuals or have entirely
equal domains, but diversion in the world's domains will sometimes be useful and
unavoidable.

@@

Finally, logic expressions in \acurt{} do not typically carry any kind of type,
since \prol{} does not employ a type system. The grammar rules of \curt{} (which
\emph{do} have \stress{syntactic} types) ensure a correct application of
unsaturated formulae. They belong in the domain of first-order predicate logic,
with two enhancements, introduced in section \ref{sec:protoq} and \ref{sec:epistemic}

\subsection{On the Proper Treatment of Questions}

Question semantics are an interesting subproblem of modern research on natural
language semantics. While traditional approaches based on \abbr{IL} and newer
accounts also employ plain extensional logic, % todo: which ones 
the treatment
of questions seems to demand that we introduce a new kind of logic formalism.
But as interesting topics go, the semantics of questions has attracted many
different views on this issue.

This section is going to elaborate on some of the approaches in greater detail
and also explain \acurt's notion of a question in particular. Note that the
latter differs from the literature, but most differences are superficial.

\subsubsection{The Question of Domain}

Often question semantics has been reduced to a problem of
pragmatics, treating direct interrogatives as nothing more than a
paraphrase of embedded interrogatives.\cite{tichy}
even goes as far as claiming equality in semantic (truth) value between
indicatives and interrogatives. The benefit of such an approach would be the
strict reliance on traditional analyses, where interrogatives would only infer a
different `attitude' in the speaker's utterance. Putting this attitude into
context and making it yield the correct results would then be left to
pragmatics.

The issue is ultimately whether the answering of questions belongs in the realm
of pragmatics or semantics. Tichy's approach has been challenged and proven
unfeasible by \cite{gs:q}, who argue for a semantic interpretation of questions.
They also propose certain adequacy criteria for theories of the semantics of
questions in order to be able to compare different approaches made in the
literature and also to defend their own solution. Their analysis is based on the
introduction of another type of logical expression, an \term{interrogative}.
In contrast to indicatives, interrogatives do not yield a truth value, but
instead pose a \emph{question}. They will be formally introduced in section
\ref{sec:protoq}.

It is important to note that \cite{karttunen:1977} attempts a treatment of
questions in a similar manner, but also \emph{denies} a difference in logical
type for questions as such. According to his hypothesis, direct questions are a
syntactic variant of indirect (embedded) questions and yes/no interrogatives are
a ``syntactically degenerate variant'' of constituent interrogatives.

\begin{quotes}A direct question can be treated as semantically equivalent to a
certain kind of declarative sentence containing the corresponding indirect
question embedded under  a suitable `performative' verb.\end{quotes}

Using this strategy, Karttunen escapes the treatment of direct  questions and
thus gives an approach for indirect (i.e. \emph{embedded}) questions  only.
In this respect, his analysis is somewhat similar to the one presented in
\cite{gs:sawhq}, but the latter goes on to challenge certain aspects of his
hypothesis.

@@

Their notion of questions will be extended for \acurt{} in order to account for
yes/no interrogatives and larger sets of answers made possible by using model
which allow for possible worlds.

Constituent interrogatives also differ in what they ask for. One can assume the
existence of at least three different types:

\begin{enumerate}
  \item NP-questions ask for a noun phrase, e.g. \example{Who shot Vincent?
  Butch (shot Vincent.)} or even more then one noun phrase at the same time, \example{Who had
  what in the hash bar?\footnote{Multiple constituent interrogatives pose a
  somewhat greater challenge, since they come with an inherent ambiguity, the
  so-called \term{Baker Ambiguity}, as presented in \cite{baker:1970}.
  Such ambiguities will not be concerned here.}}
  \item PP-questions ask for prepositional constituent, e.g. \example{Where did
  Jules stay? In Amsterdam.}
  \item VP-questions ask for an action performed by an individual, e.g.
  \example{What did Butch do? He shot Vincent.}
  \item Temporal questions ask for the date of a certain event, e.g.
  \example{When did Butch shoot Vincent? When he entered the toilet.}
  \item etc\ldots
\end{enumerate}

This classification of interrogatives can go on for quite some more entries,
including local, modal, causal and all sorts of other questions. Out of all
those categories, only NP-questions will be supported, with an option to
integrate other types of interrogatives at a later stage.

Another important distinction in the definition of questions, is the definition
of \term{answerhood}, i.e. what constitutes a valid answer to a given
interrogative. The ground-laying work by \cite{hamblin:q} states that in knowing
all possible answers to a question, one also knows the question itself. The
notion of answerhood is going to be elaborated on in section \ref{sec:answerhood}

One thing remains clear though, no matter what the notion of a question or
finding an answer to it is; upon
finding an answer, said answer has to be \emph{realized} in natural language
using text or speech synthesis. When formulating an answer, many pragmatic
considerations have to be taken into account.
However, as the uttering of an embedded question does not prompt an
answer\footnote{Unless, of course, the question is uttered as an
\term{imperative} or \term{performative}-- but this is an entirely different
class of problems that is not discussed here.}, the present
work will not directly concern itself with a solution to this problem.

@@

\subsubsection{The Semantics of Interrogatives: The ? Operator}\label{sec:protoq}

\pn{Helpful Curt}, the most advanced version of \curt{} \cite{blackburnbos:cl1}
present, only addresses the problem of direct constituent interrogatives and
does neither treat yes/no interrogatives nor embedded questions. The
interrogative operator, as it will be introduced shortly, is designed to cover
the semantics of both types of questions simultaneously. It takes a proposition
(a formula yielding a truth value as a result) but gives back a set of
propositions. Computation on more than one proposition in parallel will be
introduced into the \curt{} systems in section \ref{sec:possworlds}.  \acurt{}
represents embedded questions by embedding a special construct, a
\term{proto-question}, under verbs of certain types (which will be discussed
later on). Since the yes/no interrogative is considered a special case of the
constituent interrogative the semantic formulae produced by the grammatical
back end of \acurt{} do not differ for the two cases.

\begin{definition}\label{def:qop}
  \stress{The Question Operator}
  \begin{enumerate}
    \item If $\varphi$ is a proposition in which the free variables
    $x_1,\ldots,x_n$ for arbitrary values of $n \geq 1$ occur,
    $?\pair{x_1,\ldots,x_n}.\varphi$ denotes a \term{question}, associated with
    a constituent interrogative.
    \item If $\varphi$ does not contain any free variables, $?.\varphi$ is
    considered a question associated wit a yes/no interrogative.
    \item \label{def:outermost}The question operator does \emph{not combine} with any other operator.
  \end{enumerate}
\end{definition}

@@

Note that the last clause of definition \ref{def:qop} enforces the question operator to
always be the last operator of a formula. This constraint will be overridden
during parsing of questions to produce forms that do have nested questions.
When calculating the models and readings of a formula, however, these operators
will be resolved first. This is done in order to increase the efficiency during
parsing, since the syntactic parser will sometimes have to backtrack.

As was already mentioned, \pn{Helpful Curt} has a certain albeit limited
support for evaluating NP constituent interrogatives. Blackburn and Bos adopt a
sort of pseudo-notation of the question operator in what they call an
extra-logical form which does not itself yield a truth value, but is
internally evaluated in a special case.

They treat a question as a \emph{quantification
over a certain individual $X$, for which $R$ and $S$ must hold.} Thus a parse of
a sentence like \ref{ex:bbqueexa} and \ref{ex:bbqueexb} would yield the representation given in
\ref{ex:bbqueparsea} and \ref{ex:bbqueparseb}.

\ex. \a. \label{ex:bbqueexa} Which boxer likes Jody?
\b. \label{ex:bbqueparsea} \code{que(X, boxer(X), like(X, jody))}

\ex. \a. \label{ex:bbqueexb} Who shot Vincent?
\b. \label{ex:bbqueparseb} \code{que(X, person(X), shoot(X, vincent))}

The aforementioned special case in \pn{Helpful Curt} only evaluates a question
against the extensional model which is used as a representation of \curt's state
and knowledge. In this it differs substantially from the approaches made by
Groenendijk and Stokhof and also later approaches by \cite{g:is}. \acurt{}
bridges the gap between both worlds.

\subsubsection{What Answers the Question?}\label{sec:answerhood}

Knowing the structure of an interrogative, the next step is to evaluate the
question's \stress{value}, namely the propositions it generates. This will
invariably bring up the question of the definition of \term{answerhood}, viz.
the relation of the question to its answer or answers with respect to the
current discourse. It is important to note that an answer other than `no'
`nobody' cannot occur without a certain discourse or background knowledge already being present
for the answer to be evaluated against.

@@

For yes/no interrogatives this means, that if the question is $?.\varphi$, the
answer two it is either $\varphi$ or $\neg\varphi$. For constituent
interrogatives, this question is not so easily answered.  There is one central
issue to be addressed first: the issue of the definition of answerhood: should
the set of answers be an \emph{exhaustive} set of answers, including the
negative ones, or just the set of possible \stress{true} answers?

The following situation might serve as an example: Suppose Mia and Vincent both
play air guitar. Butch and Marcellus do not play air guitar. When answering the
question \example{Who plays air guitar?}, one can observe three degrees of
exhaustiveness, according to \cite[section 1.5]{gs:sawhq}. Clearly, it does not
suffice to just say \example{Vincent.} or \example{Mia.}, since that would still
not account for all possible answers to the question, and that both
\example{Butch.} and \example{Marcellus.} would be outright wrong. In the
analysis of \cite{karttunen:1977}, every set of propositions that is a superset
of the set of propositions containing \example{Mia plays air guitar.} and
\example{Vincent plays air guitar.} would be a valid answer. This is the lowest
degree of exhaustiveness, and accounts for all \emph{true} answers. So, if
Yolanda believed to not only know of Vincent and Mia that they played air
guitar, but also of Butch, she would still know all possible true answers, and,
according to this analysis, a correct answer.

Groenendijk and Stokhof argue for two more degrees of exhaustiveness. The first
one would require the answer not only to contain all possible \emph{true}
propositions, but also to \emph{not contain any \stress{false} propositions}. So
in this case, if Yolanda believed Butch played air guitar, she would not know
the true answer.

The second and strongest degree of exhaustiveness would
require any set of answers not only to contain the set of propositions that
hold, but also the set of propositions that \emph{do not hold}. In this case, for
the sentence \example{Yolanda knows who plays air guitar.} to be true, the
sentence \example{Yolanda knows who does not play air guitar.} would
automatically follow.

There even more ways of categorizing the number of answers to a question. So far
the so called \term{mention-all} case has been discussed, but when considering
a sentence like \example{John knows where to buy a Chinese newspaper.} (example
due to \cite{rooy:qar}), it is clear that only one or two out of all possible
propositions for \example{Where to buy a Chinese newspaper?} will be sufficient
for John to know in order for him to know where to actually buy one. This has
been regarded as the \term{mention-some} reading.

Even after having settled on the \emph{number} and \emph{value} of propositions
that have to serve as an answer to a question, one still might be concerned with
their kind. Taken the previously given example, if Butch and Marcellus were
all boxers in this situation, would not an answer such as \example{Every boxer.}
suffice? \cite{rooy:qar} elaborates on the many aspects of different analyses
regarding the relevance of answers. Suffice to say, \acurt{} is only content
with \emph{maximally specific} resolutions to answers.

Moreover, as \acurt{} will only have to deal with answers to questions
internally, it will also only regard maximally exhaustive interpretations of
answers.
Correctly answering a direct question with respect to relevance and other
pragmatic factors is a task left as a task for turning \curt{} into a complete
dialogue system more potent of dealing with queries to its knowledge.

\subsection{Embedded Questions}\label{sec:embed}

After having agreed on a treatment of interrogatives, the next step is to
construct a framework for embedding the result under certain verbs. The main
issues of this section will thus be focused on providing a classification of
embedding verbs and elaborating on the details of the relationship between
embedded and embedding constituents of an utterance.

As was already mentioned, clause \ref{def:outermost} of definition \ref{def:qop}
will be overridden when treating embedded interrogatives. The following example
illustrates a syntactic parse of an embedded interrogative:

\ex. \label{ex:whembed}
\a. Mia knows who snorts.\footnotesize
\b.  \Tree
[.$t$\\$know(?X_1.snort(X_1))(Mia)\wedge ?X_1.snort(X_1)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
[.$IV$\\$\lambda X.know(?X_1.snort(X_1))(X)\wedge ?X_1.snort(X_1)$ [.{$IV/\bar{t}$} 
{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) \wedge Y ](S) ]$\\knows } ]
[.{$\bar{t}$\\$?X_1.snort(X_1)$ } [.$\bar{t}/t$ ? ] [.{ $t$\\$snort(X_1)$ }
[.$T$ { $X_1$\\who } ]
[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{4cm} ]\normalsize
\b. $know(?X_1.snort(X_1))(Mia)\wedge ?X_1.snort(X_1)$\label{ex:formula}

@@

The reason for $?$ to be embeddable within more complex formulae is merely one
of convenience, since one could transform $?X.\varphi(X)$ to mean

\[\forall w \in W_M [ \forall x \in D_w [ \varphi(w)(x) = 1 ] ]\]

assuming $W_M$ is the set of possible worlds in a model $M$, but this would
require modifications to the current way of evaluating predicates with \curt{}
as well as the definition of a quantification over possible worlds, which is
currently not supported and would require a special treatment. Furthermore, in
the case that the explicit formula does not yield any results, the question
should return its correct answer

\[ \neg\exists x [\varphi(x)] \]

Instead, before calling its inferencing tools \acurt{} will look out only for
the question operator itself and take appropriate measures.

@@

This would again demand explicit quantification over worlds and a different
treatment of predicates. Instead, it can be treated with the same procedure as
before when using an embedded question operator.

\subsubsection{Types of Facts, Assertions and Interrogatives, Towards an
Ontology of Embedding}\label{sec:ontology}

@@

\subsubsection{Modeling Knowledge}\label{sec:epistemic}

>>>>

\documentclass[notitlepage,twoside,a4paper]{scrreprt}

@@

\usepackage{fontspec}
\usepackage[xetex,colorlinks]{hyperref}
\begin{document}
\lstset{language=Prolog,numbers=left,numberstyle=\tiny,numbersep=5pt,stepnumber=2,basicstyle=\ttfamily,identifierstyle=\textbf,keywordstyle=\underbar}
\renewcommand{\thepage}{\roman{page}}
\begin{titlepage}
  \maketitle
  \thispagestyle{empty}
  \begin{center}
    \vskip 4em
    Thesis submitted for the Degree of Bachelor of Arts\\

    \paragraph{Supervisor:} Dr. Frank Richter
    \vskip 6em
    Seminar für Sprachwissenschaft\\
    Eberhard-Karls-Universität\\
    Tübingen
  \end{center}
\end{titlepage}
\mbox{}
\newpage
\tableofcontents
\newpage
\SpruechleAufsagen{Aleksandar L. Dimitrov}
\thispagestyle{plain}
\newpage
\renewcommand{\thepage}{\arabic{page}}
\setcounter{page}{1}
\chapter{Introduction}

@@

\chapter{Theory}\label{sec:theory}

@@

\section{Formalism: Models}\label{sec:formal}

@@

As was already mentioned in \ref{sec:curt}, \curt{} is based on the use of
external inferencing tools. These tools are only capable of treating first order
logic – an extensional framework. \prol{} on the other hand, is a language that
can very well make use of higher order formalisms for calculation\footnote{The
two most popular predicates in \prol{} used for implementing higher-order logic
are \code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog}
which provides an implementation of the former, but not of the latter. In
addition \pn{SWI-Prolog} also defines several higher order aggregate functions,
including \code{maplist/4}, which will be used later on. See
\cite{naish:prolhio} for a discussion of higher order functional programming in
\prol.}, but as was already mentioned, the implementation of separate
inferencing mechanisms for \acurt{} will be avoided. While there exist
inferencing tools for modal logic\footnote{ The interested reader is referred to
\pn{Molle} (\url{http://molle.sf.net}), a modal theorem prover that comes with a
graphical user interface which allows interactive investigation of the
constructed indices.}, a restructuring of the \curt{} system to take advantage
of them would be beyond the scope of this thesis.  It was therefore decided to
avoid modal logic as much as possible and implement a pragmatic approach to
representing indices in order to increase the expressive power of the system.

Most research on the logic of questions (sometimes also referred to as eroteric
logic) suggests that the treatment of interrogatives demands its own formalism.
\acurt{} employs a formalism that departs from both \abbr{IL} and \abbr{Ty2} in
some ways. Most notably, the notion of a \term{model} is changed as follows.

@@

One of the more important differences to the `traditional' \abbr{Ty2} notion
of a model, is that it was decided to assign every world a unique domain of
discourse referents or \term{individuals}, instead of having one global set of
individuals. This follows the definition of
\cite{gs:q}, and is generally referred to as \term{partitioning of the logical
space}. \cite{tenCate} propose a
quantification over the assignment function $g$ in order to generalize a
partitioned logical space into a more standard \term{Ty2} model, however, this
would also demand the introduction of an set of an interpretation functions $I$
to assign elements of $W$ an extension, which is not necessary here because
every possible world is, in fact, an extensional model.

Note that there is no requirement $\pair{i,\pair{D,F}}, \pair{j,\pair{D',F'}}
\in M \wedge D\neq D'$, and most of the time, a model's possible worlds will
indeed share many of their individuals or have entirely equal domains, but
diversion in the world's domains will sometimes be useful and unavoidable.
%fixme: really?

@@

Finally, logic expressions in \acurt{} do not typically carry any kind of type,
since \prol{} does not employ a type system. The grammar rules of \curt{} (which
\emph{do} have \stress{syntactic} types) ensure a correct application of
unsaturated formulae. These formulae are for the most part expressions of \pn{FOL}
with two enhancements, introduced in section \ref{sec:protoq} and \ref{sec:epistemic}

\section{On the Proper Treatment of Questions}

The analysis of the meaning of questions poses an interesting task to modern
research on natural language semantics. It has so far attracted many different
theories on how to represent and compute with them.
But as interesting topics go, the semantics of questions
has attracted many different views on this issue.  It is important to first
settle on what constitutes and characterizes a \emph{direct} question before one
can turn to the treatment of \emph{embedded} questions and propositions. Most
literature seems to agree that the latter is but a special case of the former,
even though sometimes the notion of a direct question is abstracted to a
\term{proto-question}.

\subsection{The Question of Domain}

Often question semantics has been reduced to a problem of pragmatics, treating
direct interrogatives as nothing more than a paraphrase of embedded
interrogatives.\cite{tichy} even goes as far as claiming equality in semantic
(truth) value between indicatives and (embedded) interrogatives. His analysis
permits omitting a new the introduction of a new phrasal category into the
logical formalism that would otherwise be needed to account for questions.
Interrogative sentences then merely infer a different `attitude' in the
speaker's utterance. Putting this attitude into context and making it yield the
correct results is then left to pragmatics, which treats them as interrogative
acts in the setting of a dialogue.

The issue is ultimately whether the answering of questions belongs in the realm
of pragmatics or semantics. Tichy's approach, along with other, similar ones,
has been challenged by \cite{gs:q}, who argue for a treatment of interrogatives
on a semantic level, conceding a \emph{meaning} to interrogatives.  They also
propose certain adequacy criteria for theories of the semantics of questions in
order to be able to compare different approaches made in the literature and also
to defend their own solution. Their analysis is based on the introduction of
another type of logical expression, a \term{question}, which embeds a
proposition (with possibly one or more free variables) under a question operator
\term{?}.  In contrast to indicatives, questions do not yield a truth value, but
are defined by means of their \term{possible answers}.  They are formally
introduced in section \ref{sec:protoq}.

%\cite{karttunen:1977} attempts a treatment of
%questions in a similar manner, but also \emph{denies} a difference in logical
%type for questions as such. According to his hypothesis, direct questions are a
%syntactic variant of indirect (embedded) questions and yes/no interrogatives are
%a ``syntactically degenerate variant'' of constituent interrogatives.

%\begin{quotes}A direct question can be treated as semantically equivalent to a
%certain kind of declarative sentence containing the corresponding indirect
%question embedded under  a suitable `performative' verb.\end{quotes}

%Using this strategy, Karttunen escapes the treatment of direct  questions and
%thus gives an approach for indirect (i.e. \emph{embedded}) questions  only.
%In this respect, his analysis is somewhat similar to the one presented in
%\cite{gs:sawhq}, but the latter goes on to challenge certain aspects of his
%hypothesis.

@@

Their notion of questions was modified for \acurt{} in order to account for
yes/no interrogatives and larger sets of answers made possible by using a model
representation which allows for possible worlds.

\subsubsection{The Grammatical Domain of \acurt}

Constituent interrogatives further differ in \emph{what} they ask for. One can
assume the existence of at least three different types:

\begin{enumerate}
  \item NP-questions ask for a noun phrase, e.g. \example{Who shot Vincent?
  Butch (shot Vincent.)} or even more then one noun phrase at the same time, \example{Who had
  what in the hash bar?}\footnote{Multiple constituent interrogatives pose a
  somewhat greater challenge, since they come with an inherent ambiguity, the
  so-called \term{Baker Ambiguity}, as presented in \cite{baker:1970}.
  Such ambiguities will not be concerned here.}
  \item PP-questions ask for prepositional constituent, e.g. \example{Where did
  Jules stay? In Amsterdam.}
  \item VP-questions ask for an action performed by an individual, e.g.
  \example{What did Butch do? He shot Vincent.}
  \item Temporal questions ask for the date of a certain event, e.g.
  \example{When did Butch shoot Vincent? When he entered the toilet.}
  \item etc\ldots
\end{enumerate}

This classification of interrogatives can go on for quite some more entries,
including local, modal, causal and many sorts of other questions. Out of all
those categories, only NP-questions are be supported by \acurt{}, with an option to
integrate other types of interrogatives at a later stage.

\subsubsection{Pragmatic Concerns}

Another important distinction in the definition of questions, is the definition
of \term{answerhood}, i.e. what constitutes a valid answer to a given
interrogative. The different positions on this topic are elaborated on in
section \ref{sec:answerhood}. Regardless of the concept of an optimal answer, it
is clear that upon finding an answer, said answer has to be \emph{realized} in
natural language using text or speech synthesis. When formulating an answer,
many pragmatic considerations have to be taken into account.  However, as the
uttering of an embedded question does not prompt an answer\footnote{Unless, of
course, the question is uttered as an \term{imperative} -- but this is an
entirely different class of problems that is not discussed here.}, the present
work will not directly concern itself with a solution to this problem.

@@

\subsection{The Semantics of Interrogatives: The ? Operator}\label{sec:protoq}

\pn{Helpful Curt}, the most advanced version of \curt{} \cite{blackburnbos:cl1}
present, only addresses the problem of direct constituent interrogatives and
does not treat yes/no interrogatives or embedded questions. The interrogative
operator, as it will be introduced shortly, is designed to cover the semantics
of both types of questions simultaneously. It takes a proposition (a formula
yielding a truth value as a result) and yields a set of propositions. \acurt{}
represents embedded questions by embedding a special construct, a
\term{proto-question}, under verbs of certain types (discussed in section
\ref{sec:ontology}). Since the yes/no interrogative is considered a special case of the
constituent interrogative the semantic formulae produced by the grammatical back
end of \acurt{} show no difference in structure for the two cases.

\begin{definition}\label{def:qop}
  \stress{The Question Operator}, following \cite{gs:q}
  \begin{enumerate}
    \item If $\varphi$ is a proposition in which the free variables
    $x_1,\ldots,x_n$ for arbitrary values of $n \geq 1$ occur,
    $?\pair{x_1,\ldots,x_n}.\varphi$ denotes a \term{question}, associated with
    a constituent interrogative.
    \item If $\varphi$ does not contain any free variables, $?.\varphi$ is
    considered a question associated wit a yes/no interrogative.
    \item * \label{def:outermost}The question operator does \emph{not combine} with any other operator.
  \end{enumerate}
\end{definition}

@@

Note that the last clause of definition \ref{def:qop} enforces the question
operator to always be the last operator of a formula. This constraint will be
overridden during parsing of questions to produce forms that do have nested
questions.  When calculating the models and readings of a formula, however,
these operators will be resolved first. This is done in order to increase the
efficiency during parsing, since the syntactic parser will sometimes have to
backtrack. In fact, the generation of \term{logical forms} and their
interpretation are always two distinct steps in \acurt.

As was already mentioned, \pn{Helpful Curt} has a certain, albeit limited
support for evaluating NP constituent interrogatives. Blackburn and Bos adopt a
sort of pseudo-notation of the question operator in what they call an
extra-logical form. It does not yield a truth value, but is internally evaluated
in a special case.

\curt{} sees a question as a \emph{quantification over a certain individual
$X$}, for which two conditions, $R$ and $S$, must hold. \cite{bosgabdsil} call
$R$ and $S$ the \term{domain} and \term{body} of the question. The domain of a
question is usually ontological information, posing a restriction on the
quantified variable. It is selected according to the \wh-pronoun that triggered
it. So \example{who} prompts the domain to contain $person(x)$, whereas
\example{which}, used as a quantifier in a phrase as \example{which boxer}, would
impose the restriction $boxer(x)$. \ref{ex:bbqueexa} and \ref{ex:bbqueexb} give
an example of how \curt{} parses constituent interrogatives.

\ex.\label{ex:bbqueexa}
\a. Which boxer likes Jody?
\b. $?x.boxer(x) \wedge like(x,jody)$
\b. \label{ex:bbqueparsea} \code{que(X, boxer(X), like(X, jody))}

\ex.\label{ex:bbqueexb}
\a. Who shot Vincent?
\b. $?x.person(x) \wedge shoot(x, vincent)$
\b. \label{ex:bbqueparseb} \code{que(X, person(X), shoot(X, vincent))}

Similarly, yes/no interrogatives, which are new to the \curt{} system, are
parsed using the same sort of operator:\footnote{The actual representation in
the code will differ a little in order to work around a bug in the system.}

\ex.
\a. (whether) Vincent snorts.
\b. $?.snort(Vincent)$
\b. \lstinline!que([], [], snort(vincent))!

\ex. \a. (whether) Mia or Jody dances.
\b. $?.dance(mia) \vee dance(jody)$
\b. \lstinline!que([], [], or(dance(mia),dance(jody)))!

The aforementioned special case in \pn{Helpful Curt} only evaluates a question
against the extensional model which is used as a representation of \curt's state
and knowledge. In this it differs substantially from the approaches made by
Groenendijk and Stokhof and also later approaches by \cite{g:is}. \acurt{}
aims to bridge the gap between both worlds.

\subsection{What Answers the Question?}\label{sec:answerhood}

Knowing the structure of an interrogative, the next step is to evaluate the
question's \stress{value}, namely the propositions it generates. This will
invariably bring up the question of the definition of \term{answerhood}, viz.
the relation of the question to its answer or answers with respect to the
current discourse. Note that an answer other than `no' or
`nobody' cannot occur without a certain discourse or background knowledge already being present
for the answer to be evaluated.

@@

For yes/no interrogatives this means, that if the question is $?.\varphi$, the
answer to it is either $\varphi$ or $\neg\varphi$. For constituent
interrogatives, this question is not so easily answered.  There is one central
issue to be addressed first: should the set of answers be an \emph{exhaustive}
set of answers, including the negative ones, or just the set of possible
\emph{true} answers?

The following situation might serve as an example: Suppose Mia and Vincent both
play air guitar. Butch and Marsellus do not play air guitar. When answering the
question \example{Who plays air guitar?}, one can observe three degrees of
exhaustiveness, according to \cite[section 1.5]{gs:sawhq}. Clearly, it does not
suffice to just say \example{Vincent.} or \example{Mia.}, since that would still
not account for all possible answers to the question, and that both
\example{Butch.} and \example{Marcellus.} would be outright wrong. In the
analysis of \cite{karttunen:1977}, every set of propositions that is a superset
of the set of propositions containing \example{Mia plays air guitar.} and
\example{Vincent plays air guitar.} would be a valid answer. This is the lowest
degree of exhaustiveness, and accounts for all \emph{true} answers. So, if
Yolanda believed to not only know of Vincent and Mia that they played air
guitar, but also of Butch, she would still know all possible true answers, and,
according to this analysis, have a correct answer.

\cite{gs:sqpa} argue for two more degrees of exhaustiveness. The first
one requires the answer not only to contain all possible \emph{true}
propositions, but also to \emph{not contain any \stress{false} propositions}. So
in this case, if Yolanda believed Butch played air guitar, she would not know
the true answer.

\footnotetext{
There even more ways of categorizing the number of answers to a question. So far
the so called \term{mention-all} case has been discussed, but when considering
a sentence like \example{John knows where to buy a Chinese newspaper.} (example
due to \cite{rooy:qar}), it is clear that only one or two out of all possible
propositions for \example{Where to buy a Chinese newspaper?} will be sufficient
for John to know in order for him to know where to actually buy one. This has
been regarded as the \term{mention-some} reading.}

The second and strongest degree of exhaustiveness 
requires any set of answers not only to contain the set of propositions that
hold, but also the set of propositions that \emph{do not hold}. In this case, for
the sentence \example{Yolanda knows who plays air guitar.} to be true, the
sentence \example{Yolanda knows who does not play air guitar.} would
automatically follow.

\subsubsection{Combinatorial Problems}

Thus the maximum degree of exhaustiveness bears several problems. The most
obvious one is that of a considerable \emph{combinatoric explosion}\footnotemark. In the
previous example, containing only four individuals, the question
$?x.play\_air\_guitar(x)$ evaluates to a total set of $2^4$ propositions:
\example{Nobody plays air guitar.}, \example{Only Butch plays air guitar},
\example{Only Mia plays air guitar},\ldots{}, \example{Only Butch and Mia play
air guitar},\ldots{}, \example{Everybody plays air guitar}. Generally, every
question will evaluate to $2^{|D|}$ propositions. As \cite{bosgabdsil} show, to
find a suitable answer to a given partition of logical space (i.e. a
\emph{question}), the logical inference tools would have to perform consistency
checks on the union \emph{every single} possible answer with the discourse.

\footnotetext{
There is at least one more problem with this analysis although \cite{gs:q} hint
at an easy solution for this. Suppose Yolanda does not
know Butch, but she knows Mia and Vincent, and that they play air guitar. But since
she does not know about Butch, she has no chance of knowing a \emph{most
specific} answer. Groenendijk and Stokhof's analysis thus requires every
individual to have exhaustive knowledge about the discourse domain.}

\cite{bosgabdsil} go on to propose a method for taming the computational
complexity of Groenendijk and Stokhof's analysis, but they do so at the cost of
a strong account for exhaustivity. Their system reduces the size of the
partition to essentially four propositions: \example{Everybody plays air
guitar.}, \example{Somebody plays air guitar.}, \example{Somebody does not play
air guitar.}, \example{Nobody plays air guitar.} Now the inferencing task has
shrunk down considerably, but it does not even require \emph{every true answer}
to be known in the case that the body of the question does not hold for at least
one individual, which will, for non-trivial domains, almost always be the case.

As \acurt{} will only have to deal with answers to questions
internally, it will also only regard maximally exhaustive interpretations of
answers.

Furthermore, correctly answering a direct question with respect to relevance and other
pragmatic factors is a task left as a task for turning \curt{} into a complete
dialogue system more potent of dealing with queries to its knowledge.

\section{Embedded Questions}\label{sec:embed}

After having agreed on a treatment of interrogatives, the next step is to
construct a framework for embedding the result under certain verbs.  This
section provides a classification of embedding verbs and elaborates on the
details of the relationship between embedded and embedding constituents of an
utterance.

\subsection{Syntax}
As was already mentioned, clause \ref{def:outermost} of definition \ref{def:qop}
will be overridden when treating embedded interrogatives. The following example
illustrates a syntactic parse of an embedded interrogative:

\ex. \label{ex:whembed}
\a. Mia knows who snorts.\footnotesize
\b.  \Tree
[.$t$\\$know(?X_1.snort(X_1))(Mia)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
[.$IV$\\$\lambda X.know(?X_1.snort(X_1))(X)$ [.{$IV/\bar{t}$} 
{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X)](S) ]$\\knows } ]
[.{$\bar{t}$\\$?X_1.snort(X_1)$ } [.$\bar{t}/t$ ? ] [.{ $t$\\$snort(X_1)$ }
[.$T$ { $X_1$\\who } ]
[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{5cm} ]\normalsize
\b. $know(?X_1.snort(X_1))(Mia)$\label{ex:formula}

@@

The reason for $?$ to be embeddable within more complex formulae is not merely one
of convenience, since it cannot be evaluated without access to a particular
world's set of individuals $D$. This would mix semantic interpretation with
the parsing process and such side-effects were not deemed desirable.
Instead, before calling its inferencing tools \acurt{} will look out only for
the question operator itself and take appropriate measures.

@@

Obviously, the partitions for yes/no interrogatives is much smaller than it is
for constituent interrogatives and it does not need access to a particular model
in order to form its partition. Yet this is not reason enough to give up the
more general approach already described.

\subsection{Types of Facts, Assertions and Interrogatives, Towards an
Ontology of Embedding}\label{sec:ontology}

@@

\subsection{Modeling Knowledge}\label{sec:epistemic}
